---
title: "searchK and selectModel Methods"
author: "A.Pedraza"
output: html_document
date: "2024-12-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
#library(magrittr)
library(dplyr)
library(tidyr)
library(kableExtra)
library(tm) # Load the tm package
library(stm)
library(wordcloud)
```

<br>

**Summary of data frame**

- Data used from csv file: **final_df_condition.csv**

- Search into Pub Med included: Case Report and Clinical Trials of the following disease:
  
    + Cancer
    + Viruses
    + Parkinson's Disease
    + Cardiovascular Disease

<br>

```{r}
df<- read.csv("final_df_condition.csv")
print(summary(df))

```

<br>

**Dataframe example**

```{r, warning=FALSE}
kbl(head(df)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

<br>

<br>

### Histogram of Years Represented

  - The number of years is slightly off balance. It was difficult to get articles with abstracts in the later years.
  
<br>

```{r, warning=FALSE}

df %>%
  ggplot( aes(x=Year, fill=Year)) +
    geom_histogram(bins=15, fill='skyblue', color='#69b3a2', alpha=.6, width=.4) + scale_x_continuous(breaks=seq(2010,2024,1))+
    #scale_fill_manual(values=c("#69b3a2", "#404080")) +
    theme_bw() + ggtitle("Number of Articles in Dataset per Year") + xlab("Year of Publication") + 
  ylab("Number of Articles")


```

<br>

- A balanced number of medical conditions represented in the data set.

```{r}
df%>% ggplot( aes(x=Condition, fill=Condition)) + geom_bar(fill='orange', color='#69b3a2', alpha=.6, width=.4) + ggtitle("Number of Abstracts per Condition")

```
<br>

<br>


**Processing the data and text**

  - Used textProcessor() from stm library to process the metadata and the documents. I chose 'stem = FALSE' because I did not want complicated medical terms to be misunderstood in the analysis.
  
  - The metadata is: 'Condition', 'Year', 'Journal'
  
  - The documents is the  'Abstract' column. 
```{r}
# Process the text

documents <- df$Abstract
journal <- df$Journal
condition <- df$Condition
year <- df$Year
#title <- df$Title
metadata <- data.frame(journal, year, condition)

set.seed(1234)
processed <- textProcessor(df$Abstract, metadata = metadata, stem=FALSE)
out_cond <- prepDocuments(processed$documents, processed$vocab, processed$meta, lower.thresh = 25)

docs_cond <- out_cond$documents
vocab_cond <- out_cond$vocab
meta_cond <- out_cond$meta

```

br>


#### SearchK()

  - Use searchK() to find the best number of Topics (value of K) to use in model search.

```{r}
k_result_cond <- searchK(documents = docs_cond, vocab = vocab_cond, prevalence = ~ year +  journal, K = seq(3, 10) , data = meta_cond, max.em.its = 200, gamma.prior = 'L1')

```


<br>

<br>



#### Plot of Search_K result
***

<br>

##### Diagnostic Plots

  - High semantic coherence for K = 5, but slightly higher Residuals and slightly low held-out likelihood.
  
<br>

```{r}
# Plots of ksearch results
plot(k_result_cond)
```


<br>

<br>

##### Semantic Coherence and Exclusivity Plot and Composite Score
***

<br>

**definitions**

  - Semantic Coherence: When the most probable words to appear together in a given topic frequently occur together, this score is high. So we are looking for a high score in semantic coherence because that will tell us that the words of a given topic occur together with high frequency.
  
  - Exclusivity: tells us how strongly a set of words in a Topic is associated with that Topic because it is exclusive, most likely only found in that Topic rather than other topics.
  
  
<br>

**Composite Score Plot** 

  - Composite Score calculation: 
  
    + Composite Score = 0.5 × Coherence + 0.5 × Exclusivity

  - We are looking for a Topic with both a high Semantic Coherence score and a Exclusivity score.

```{r}
# get semantic coherence and exclusivity scores put in dataframe
semcoh <- sapply(k_result_cond$results$semcoh, function(x) x[1])
excl <- sapply(k_result_cond$results$exclus, function(x) x[1])
k_val <- seq(3, 10)
scores_df <- data.frame(Topics = k_val, Coherence = semcoh, Exclusivity = excl)

# Composite score
composite_score <- numeric()
for (i in 1:nrow(scores_df)) {
  
  composite_score[i] <- (scores_df[i, 2] * (0.5)) + (scores_df[i, 3] * (0.5))

}

scores_df$composite_score<- composite_score
```

<br>

  - Use the plot below to select the best value for K, the number of topics to be included in model.
  
  - The value of K that has the best composite score will have a value high in both Semantic coherence and Exclusivity.
  
  - From the plot below, it is evident that **K = 3, total number of topics in the model**, has the best Composite Score. This is the number for **K** that will be used in selectModel().
  
<br>

```{r}
ggplot(scores_df, aes(x = Topics, y = composite_score, label = Topics)) +
  geom_point(colour = "green", size = 3) +
  geom_text(vjust = -0.5) +
  labs(title = "Topics vs. Composite Scores", x = "Number of Topics", y = "Composite Scores")+
  theme_minimal()+
  theme(axis.line = element_line(linewidth = 0.75),
    axis.line.x.bottom = element_line(colour = "blue"),
    axis.line.y.left = element_line(colour = "orange"))
    
```



<br>

### Select Model() with K = 3: Total Topics in Model

  - Using selectModel() with number of topics = 3 to help find the best model for our data.
  
  - Using gamma.prior = 'L1' to help improve speed and convergence.
<br>

```{r cars}
select_model_cond <- selectModel(out_cond$documents, out_cond$vocab, prevalence = ~ year +  journal, K = 3, data = out_cond$meta, runs = 30, seed =1234, max.em.its = 200, gamma.prior='L1')


```


<br>

#### **Plot Semantic Coherence and Exclusivity Scores:**
***

- After 105 iterations, the model converged.

<br>

**Plot of Semantic Coherence and Exclusivity Scores of the different models found**

<br>

- From looking at the plot below, it looks like both Model 1 and Model 5 have the best overall scores. 
```{r pressure, echo=FALSE}
plotModels(select_model_cond, pch=c(1,2,3,4,5,6), legend.position="bottomright", main = "Scores of Models Found from Search")
```
<br>

#### **Composite Scores for Exclusivity and Coherence**

<br>

- The below plot confirms that **Model 1** has the best overall exclusivity and semantic coherence scores.


```{r}
semcoh_2 <- sapply(select_model_cond$semcoh, function(x) x[1])
excl_2 <- sapply(select_model_cond$exclusivity, function(x) x[1])

k_val <- seq(1, 6)
scores_df_2 <- data.frame(Topics = k_val, Coherence = semcoh_2, Exclusivity = excl_2)
scores_df_2
```

```{r}

# Composite score
composite_score <- numeric()
for (i in 1:nrow(scores_df_2)) {
  # column 2 is coherence
  # column 3 is exclusivity
  composite_score[i] <- (scores_df_2[i, 2] * (0.5)) + (scores_df_2[i, 3] * (0.5))

}

scores_df_2$composite_score<- composite_score
scores_df_2
```


```{r}
ggplot(scores_df_2, aes(x = Topics, y = composite_score, label = Topics)) +
  geom_point(colour = "green", size = 3) +
  geom_text(vjust = -0.5) +
  labs(title = "Topics vs. Composite Scores", x = "Model Number", y = "Composite Scores")+
  theme_minimal()+
  theme(axis.line = element_line(linewidth = 0.75),
    axis.line.x.bottom = element_line(colour = "blue"),
    axis.line.y.left = element_line(colour = "orange"))

```

<br>

### Selecting Model and Printing Results

<br>

  - Selecting model #1 because it seems to have the highest overall Exclusivity and Semantic Coherence scores even though model 4 is very close to it.
  
  
```{r}

selected_model_1 <- select_model_cond$runout[[1]]

```

<br>

#### Model 1: Top 7 Words for each of the 3 Topics
***

<br>

  - You can use the summary() or labelTopics functions to see the top 7 words associated with each of the 3 Topics.

<br>

```{R}
labelTopics(selected_model_1, c(1, 2, 3))
```

<br>

### Top Abstracts for Each of the Topics and Labeling
***

  - Extract thetas from the model or use findThought() function.
  
  - With extracting thetas, use the highest theta values for each topic and its row index to find the abstract associated with that theta.
  
  - I labeled each topic using the top 7 words for each topic (both HIGHEST and FREX) as well as reading the abstract and summarizing in 4-5 words for the label. 
  

<br>


**Topic 1: Journal, Year, and Abstract**

<br>

- From reading the abstract most associated with this topic, I'm putting the following label on Topic 1:

  + Topic 1: **Protein and Cell Expression**

<br>

```{r}
# Extract the thetas from the selected model
theta_matrix <- selected_model_1$theta

# Make matrix into a data frame and change column names
theta_df_1 <- as.data.frame(theta_matrix)
colnames(theta_df_1) <- c("Topic1","Topic2", "Topic3")



# make a series of the thetas for each topic
topic1 <- theta_df_1$Topic1
topic2 <- theta_df_1$Topic2
topic3 <- theta_df_1$Topic3

```

```{r}
# Index numbers for the top theta value
list <- tail(order(topic1),5)
first = list[5]
second = list[4]
```

```{r}
# Theta values in decreasing order
theta_1 <- theta_df_1[first,1]
theta_2 <- theta_df_1[second,1]


# The Top Review most associated with Topic 1
print(paste0("First Article Theta Value:  " ,theta_1))
print(paste0("Title of Article: ", df[first,1]))
print(paste0("Journal Name: ",df[first,3]))
print(paste0("Year: ",df[first,4]))
print(paste0("Abstract 1: ", df[first,2]))
print(paste0("Second Abstact Theta Value:  " ,theta_2))
print(paste0("Abstract 2: ", df[second,2]))

```

<br>

**Topic 2:  - using findThoughts()**

<br>

- From reading the abstract most associated with this topic, I'm putting the following label on Topic 2:

  + Topic 2: **Rare Cancers and their Treatment**
  

```{r}
topic2_thoughts <- findThoughts(selected_model_1, text = documents, n=2, topics = 2)$docs[[1]]
print(paste0(topic2_thoughts[1]))

print(paste0(topic2_thoughts[2]))

```


<br>

<br>

**Topic 3:**

<br>

- From reading the abstract most associated with this topic, I'm putting the following label on Topic 3:

  + Topic 3: **Cardiovascular Disease Risks and Predictions**

```{r}
topic3_thoughts <- findThoughts(selected_model_1, text = documents, n=2, topics = 3)$docs[[1]]
print(paste0(topic3_thoughts[1]))

print(paste0(topic3_thoughts[2]))

```



<br>

#### Word Clouds
***

<br>

- Using word clouds to get a more visual look into what are the most frequent and exclusive words for each topic according to the model.

**Topic 1**

<br>

```{r, warning = FALSE}
par(mfrow = c(2, 2), mar = c(.1,.1,.1,.1))
for (topic in 1:3) {
  cloud(selected_model_1, topic = 1, scale = c(2, 0.5), max.words = 50, colors = brewer.pal(8, "Dark2"))
}
```

<br>

<br>

**Topic 2**

```{r, warning = FALSE}
par(mfrow = c(2, 2), mar = c(.1,.1,.1,.1))
for (topic in 1:3) {
  cloud(selected_model_1, topic = 2, scale = c(2, 0.5), max.words = 50, colors = brewer.pal(8, "Dark2"))
}
```

<br>

**Topic 3**

```{r, warning = FALSE}
par(mfrow = c(2, 2), mar = c(.1,.1,.1,.1))
for (topic in 1:3) {
  cloud(selected_model_1, topic = 3, scale = c(2, 0.5), max.words = 50, colors = brewer.pal(8, "Dark2"))
}
```

<br>


### Metadata Relationships with Topics
***

  + Topic 1: **Protein and Cell Expression**

  + Topic 2: **Rare Cancers and their Treatment**

  + Topic 3: **Cardiovascular Disease Risks and Predictions**

<br>

  - Topic 1, protein and cell expression, proportion in medical literature has decreased over the years whereas topic 2, cardiovascular disease risk and predictions have increased.
.

```{r}
#out_cond$meta$condition <- as.factor(out_cond$meta$condition)
out_cond$meta$year <- as.numeric(out_cond$meta$year)
prep <- estimateEffect(1:3 ~ year ,  selected_model_1, meta = meta_cond, uncertainty = "Global")
```

```{r}
plot(prep, covariate = "year", method = "continuous", topics = c(1,2), model = selected_model_1 , main = "Proportion of Topic 1 and 2 over the past 14 years")

```

<br>

**Plot**

  - The proportion of Topic 3 **Cardiovascular Disease Risks and Predictions** has increased over the years:
  
```{r}
plot(prep, covariate = "year", method = "continuous", topics = c(3), model = selected_model_1 , main = "Proportion of Topic 3 over the past 14 years")

``` 

<br>

**Plot**

  + Topic 1: **Protein and Cell Expression**

  - This topic has decreased in proportion for both cancer and virus literature.
  
  
```{r}
prep2 <- estimateEffect(c(1) ~ condition * year, selected_model_1, metadata = out_cond$meta, uncertainty = "None")

plot(prep2, covariate = "year", model = selected_model_1, method="continuous", xlab = "Years", moderator = "condition", moderator.value = "cancer", linecol = "blue", printlegend = FALSE, ylim = c(0, 1.0), main = "Proportion of Topic 1 for Cancer and Virus Publications")
plot(prep2, covariate = "year", model = selected_model_1, method="continuous", xlab = "Years", moderator = "condition", moderator.value = "virus", linecol = "pink", printlegend = FALSE, add = TRUE)
legend(x="top", legend = c("Cancer", "Virus"), lwd = 2, col = c("blue", "pink"))

```

<br>

**Plot**


  - Topic 2: **Rare Cancers and their Treatment**

  - This topic has increased a lot in cancer literature.
  
  - This topic has shown a slight increase in proportion for cardiovascular disease literature.

```{r}
prep3 <- estimateEffect(c(2) ~ condition * year, selected_model_1, metadata = out_cond$meta, uncertainty = "None")

plot(prep3, covariate = "year", model = selected_model_1, method="continuous", xlab = "Years", moderator = "condition", moderator.value = "cardiovascular disease", linecol = "green", printlegend = FALSE, ylim = c(0, 1.0), main = "Proportion of Topic 2 for CD and Cancer  Publications")
plot(prep3, covariate = "year", model = selected_model_1, method="continuous", xlab = "Years", moderator = "condition", moderator.value = "cancer", linecol = "orange", printlegend = FALSE, add = TRUE)
legend(x="top", legend = c("cardiovascular disease", "cancer"), lwd = 2, col = c("green", "orange"))

```

<br>

**Plot**


  - Topic 3: **Cardiovascular Disease Risks and Predictions**

  - This topic has decreased slightly in cardiovascular disease literature. 
  
  - This topic has increased slightly in Parkinson's disease literature. 

```{r}
prep4 <- estimateEffect(c(3) ~ condition * year, selected_model_1, metadata = out_cond$meta, uncertainty = "None")

plot(prep4, covariate = "year", model = selected_model_1, method="continuous", xlab = "Years", moderator = "condition", moderator.value = "cardiovascular disease", linecol = "green", printlegend = FALSE, ylim = c(0, 1.0), main = "Proportion of Topic 3 for PD and CD Publications")
plot(prep4, covariate = "year", model = selected_model_1, method="continuous", xlab = "Years", moderator = "condition", moderator.value = "parkinsons disease", linecol = "orange", printlegend = FALSE, add = TRUE)
legend(x="top", legend = c("cardiovascular disease", "parkinsons disease"), lwd = 2, col = c("green", "orange"))

```

<br>

<br>

**Plot**


  - Topic : **Rare Cancers and their Treatment** - with Parkinson's Disease and Cardiovascular Disease

  - This topic has increased a lot in cancer literature.
  
  - This topic has shown a slight increase in proportion for cardiovascular disease literature.

```{r}
prep5 <- estimateEffect(c(1) ~ condition * year, selected_model_1, metadata = out_cond$meta, uncertainty = "None")

plot(prep5, covariate = "year", model = selected_model_1, method="continuous", xlab = "Years", moderator = "condition", moderator.value = "parkinsons disease", linecol = "green", printlegend = FALSE, ylim = c(0, 1.0), main = "Proportion of Topic 1 for PD and cardiovascular disease Publications")
plot(prep5, covariate = "year", model = selected_model_1, method="continuous", xlab = "Years", moderator = "condition", moderator.value = "cardiovascular disease", linecol = "orange", printlegend = FALSE, add = TRUE)
legend(x="top", legend = c("parkinsons disease", "cardiovascular disease"), lwd = 2, col = c("green", "orange"))

```

<br>




